---
title: "sample 11 from DLPFC data"
output: html_document
date: "2023-11-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(SpatialPCA)
```

```{r}

#----------------
# LIBD
#----------------

dataset="LIBD"

# BiocManager::install("spatialLIBD")
library('spatialLIBD')
sce <- fetch_data(type = 'sce')
metaData = SingleCellExperiment::colData(sce)
expr = SingleCellExperiment::counts(sce)
sample_names <- paste0("sample_", unique(colData(sce)$sample_name))


i=11 ### try sample 11
sce_sub <- sce[, colData(sce)$sample_name == gsub("^sample_", "", sample_names[i])]
dim(sce_sub)
count_sub = SingleCellExperiment::counts(sce_sub)
xy_coords <- data.frame(
        x_coord = colData(sce_sub)[, c("imagecol")], 
        y_coord = -colData(sce_sub)[, c("imagerow")]
    )

```

```{r}
## Load known marker genes (from Kristen)
# load spreadsheet of known marker genes (from Kristen)
# download from https://github.com/LieberInstitute/HumanPilot/blob/master/Analysis/KRM_Layer_Markers.xlsx
library( readxl)
KRM_Layer_Markers <- read_xlsx("KRM_Layer_Markers.xlsx")
dim(KRM_Layer_Markers)
```

```{r}
marker_genes_KRM <- KRM_Layer_Markers$Gene
sum(toupper(marker_genes_KRM) %in% rowData(sce)$gene_name)
genes_markers <- data.frame(gene_name = toupper(marker_genes_KRM))
genes_markers$gene_id <- rowData(sce)$gene_id[match(genes_markers$gene_name, rowData(sce)$gene_name)]
```

```{r}
# check ground truth layer number
true_num=c()
 metaData1 = as.data.frame(metaData)
 name=unique(colData(sce)$sample_name)
for(i in 1:12){
	true_num[i] = length(na.omit(unique(metaData1$layer_guess_reordered[which(metaData1$sample_name==name[i])])))

}
true_num
#  [1] 7 7 7 7 5 5 5 5 7 7 7 7
```


```{r}
# get ground truth labels for sample
metaData1 = as.data.frame(metaData)
library(dplyr)
i = 11 
KRM_manual_layers_sub <- filter(metaData1, sample_name == gsub("^sample_", "", sample_names[i])) ##Caution, this is the dplyr filter function

dim(KRM_manual_layers_sub)

```


```{r}
i = 11
ground_truth_sub <- data.frame(truth = rep(NA, ncol(sce_sub)))
rownames(ground_truth_sub) <- colnames(sce_sub)
ground_truth_sub$truth = KRM_manual_layers_sub$layer_guess_reordered
Layer_sub = KRM_manual_layers_sub$layer_guess_reordered
sum(as.character(KRM_manual_layers_sub$barcode) == colnames(count_sub)) # yes all matched


save(Layer_sub, xy_coords, KRM_manual_layers_sub, count_sub, file = paste0("LIBD_sample",i,".RData") ) ## save sample data

```

```{r}
#load("LIBD_sample12.Rdata")
```




```{r}
library(SPARK)
library(Seurat)
library(peakRAM)
library(SpatialPCA)
library(ggplot2)

```

```{r}
xy_coords = as.matrix(xy_coords)
rownames(xy_coords) = colnames(count_sub)

LIBD = CreateSpatialPCAObject(counts=count_sub, location=xy_coords, project = "SpatialPCA",gene.type="spatial",sparkversion="sparkx",numCores_spark=1,gene.number=3000, customGenelist=NULL,min.loctions = 20, min.features=20)  ### mc.cores > 1 is not supported on Windows for mcapply. Instead of numCores_spark = 5,can use = 1 in the mean time (This step is not very time-consuming, 25 seconds at most)

```

```{r}
mem <- peakRAM({
start_time <- Sys.time()
LIBD = SpatialPCA_buildKernel(LIBD, kerneltype="gaussian", bandwidthtype="SJ",bandwidth.set.by.user=NULL)
LIBD = SpatialPCA_EstimateLoading(LIBD,fast=FALSE,SpatialPCnum=5)
LIBD = SpatialPCA_SpatialPCs(LIBD, fast=FALSE)
end_time <- Sys.time()
T = end_time - start_time
})
```

```{r}
i = 11
SpatialPCA_result = list()
SpatialPCA_result$SpatialPCs  = LIBD@SpatialPCs # extracted spatial PCs
SpatialPCA_result$location = LIBD@location
truth = KRM_manual_layers_sub$layer_guess_reordered[match(colnames(LIBD@normalized_expr),colnames(count_sub))]
SpatialPCA_result$truth = truth
```

```{r}
clusterNum=c(7,7,7,7,5,5,5,5,7,7,7,7)

```

### Compute ARI on clusters for SpatialPCA and PCA; clustering method: Walktrap

```{r}
i = 11
pred_cluster= walktrap_clustering(clusterNum[i],SpatialPCA_result$SpatialPCs,knearest=70 ) # I set knearest=70 for all 12 samples in the DLPFC data. User can try other knearest number in other datasets, e.g. knearest=round(sqrt(dim(SpatialPCA_result$SpatialPCs)[2]))
SpatialPCA_result$clusterlabel = pred_cluster
SpatialPCA_result$clusterlabel_refine=refine_cluster_10x(pred_cluster,SpatialPCA_result$location,shape="hexagon") 
```

```{r}
ind_na=which(is.na(SpatialPCA_result$truth)) # we remove will NA samples in the original annotation to calculate ARI.
library(mclust)
SpatialPCA_result$ARI = adjustedRandIndex(SpatialPCA_result$clusterlabel_refine[-ind_na],SpatialPCA_result$truth[-ind_na])
# for sample i, calculate the ARI.
```


```{r}
library(lisi)
metadata_DLPFC = data.frame("SpatialPCA" = SpatialPCA_result$clusterlabel_refine)
SpatialPCA_result$LISI <- compute_lisi(SpatialPCA_result$location, metadata_DLPFC, c('SpatialPCA'))

save(SpatialPCA_result, file = paste0("SpatialPCA_DLPFC_sample",i,"result.RData"))
```



```{r}
i=11 ## sample 11
library(ggpubr)
library(umap)

xy_coords = as.matrix(xy_coords)
rownames(xy_coords) = colnames(count_sub)
load( paste0("SpatialPCA_DLPFC_sample",i,"result.RData"))

```



```{r}

# ARI; compare estimated clusters with ground truth
library(aricode)



```


```{r}
load("LIBD_sample11_normalized.Rdata")
```

```{r}
i = 11
load( paste0("SpatialPCA_DLPFC_sample",i,"result.RData"))
ind_na=which(is.na(SpatialPCA_result$truth)) # we remove NA samples in the original annotation to calculate ARI.
library(mclust)
```

### PCA

```{r}
X0 = LIBD@normalized_expr
```

```{r}
X0_k = svd(X0)
```

```{r}
v0 = X0_k$v[,1:5]   ## number of components = 5
u0 = X0_k$u[, 1:5]
```


```{r}
clusterNum=c(7,7,7,7,5,5,5,5,7,7,7,7)
PCA_pred_cluster= walktrap_clustering(clusterNum[11],t(u0)%*%X0,knearest=70 ) 

PCA_clusterlabel_refine=refine_cluster_10x(PCA_pred_cluster,SpatialPCA_result$location,shape="hexagon")
```


```{r}
adjustedRandIndex(SpatialPCA_result$clusterlabel_refine[-ind_na],SpatialPCA_result$truth[-ind_na])
adjustedRandIndex(PCA_clusterlabel_refine[-ind_na],SpatialPCA_result$truth[-ind_na])

```



### For iterative spatial SVD method: Ouput preprocessed locality and gene expression data for downstream cross validation step

```{r}
write.csv(as.data.frame(X0), file = ".../expr_data.csv")
```

```{r}
write.csv(as.data.frame(scale(SpatialPCA_result$location)), file = ".../expr_locs.csv")
```


```{r}
load( paste0("SpatialPCA_DLPFC_sample",11,"result.RData"))
```


### Load spatial SVD outputs (computed by iterative spatial SVD): 


```{r}
IP_U = as.matrix(read.csv("IP_U.csv"))
IP_V = as.matrix(read.csv("IP_V.csv"))
```


```{r}
PCA_denoised =  (t(IP_V)%*% X0) %*% (IP_U %*% t(IP_U))
```



## ARI for iterative SVD method


```{r}
clusterNum=c(7,7,7,7,5,5,5,5,7,7,7,7)
IP_pred_cluster= walktrap_clustering(clusterNum[11],PCA_denoised,knearest=70 ) 
IP_clusterlabel_refine=refine_cluster_10x(IP_pred_cluster,SpatialPCA_result$location,shape="hexagon")

```

```{r}
ind_na=which(is.na(SpatialPCA_result$truth))
```


```{r}
adjustedRandIndex(IP_clusterlabel_refine[-ind_na],SpatialPCA_result$truth[-ind_na])
```



### We can also use Gaussian mixture for clustering, and compute corresponding ARIs.
```{r}
library(ClusterR)
```


```{r}
dat = as.data.frame(t(t(u0)%*%X0))  ## transpose 
gmm = GMM(dat, 7, dist_mode = "eucl_dist", seed_mode = "random_subset", km_iter = 10,
          em_iter = 10, verbose = F)          

# predict centroids, covariance matrix and weights
pr0 = as.factor(predict(gmm, newdata = dat))

#pr0=refine_cluster_10x(pr0,SpatialPCA_result$location,shape="hexagon")
```

#### PCA 
```{r}
adjustedRandIndex(pr0[-ind_na],SpatialPCA_result$truth[-ind_na])
```

#### Spatial SVD
```{r}
dat = as.data.frame(t(PCA_denoised))  ## transpose 
gmm = GMM(dat, 7, dist_mode = "eucl_dist", seed_mode = "random_subset", km_iter = 10,
          em_iter = 10, verbose = F)          

# predict centroids, covariance matrix and weights
pr1 = as.factor(predict(gmm, newdata = dat))

#pr1=refine_cluster_10x(pr1,SpatialPCA_result$location,shape="hexagon")
```

```{r}
adjustedRandIndex(pr1[-ind_na],SpatialPCA_result$truth[-ind_na])
```

#### Spatial PCA

```{r}
dat = as.data.frame(t(SpatialPCA_result$SpatialPCs)) ## transpose 
gmm = GMM(dat, 7, dist_mode = "eucl_dist", seed_mode = "random_subset", km_iter = 10,
          em_iter = 10, verbose = F)          

# predict centroids, covariance matrix and weights
pr2 = as.factor(predict(gmm, newdata = dat))

#pr2=refine_cluster_10x(pr2,SpatialPCA_result$location,shape="hexagon")
```

```{r}
adjustedRandIndex(pr2[-ind_na],SpatialPCA_result$truth[-ind_na])
```



## Compute the CHAOS and PAS scores:

```{r}
require(parallel)
library(pdist)
fx_CHAOS_windows = function(clusterlabel, location){
  # require(parallel)
  matched_location=location
  NAs = which(is.na(clusterlabel))
  if(length(NAs>0)){
    clusterlabel=clusterlabel[-NAs]
    matched_location = matched_location[-NAs,]
  }
  matched_location = scale(matched_location)
  dist_val = rep(0,length(unique(clusterlabel)))
  count = 0
  for(k in unique(clusterlabel)){
    count = count + 1
    location_cluster = matched_location[which(clusterlabel == k),]
    if(length(location_cluster)==2){next}
    #require(parallel)
    results = mclapply(1:dim(location_cluster)[1], fx_1NN, location_in=location_cluster,mc.cores = 1)
    dist_val[count] = sum(unlist(results))
  }
  dist_val = na.omit(dist_val)
  return(sum(dist_val)/length(clusterlabel))

}
```

```{r}
fx_1NN = function(i,location_in){
  # library(pdist)
  line_i = rep(0,dim(location_in)[1])
  line_i = pdist(location_in[i,],location_in[-i,])@dist
  return(min(line_i))
}

#' @import pdist
fx_kNN = function(i,location_in,k,cluster_in){
  #library(pdist)
  line_i = rep(0,dim(location_in)[1])
  line_i = pdist(location_in[i,],location_in[-i,])@dist
  ind = order(line_i)[1:k]
  cluster_use = cluster_in[-i]
  if(sum(cluster_use[ind] != cluster_in[i])>(k/2)){
    return(1)
  }else{
    return(0)
  }

}
```

##### CHAOS (via Walktrap): 
```{r}
fx_CHAOS_windows(PCA_clusterlabel_refine,SpatialPCA_result$location) ### PCA
```

```{r}
fx_CHAOS_windows(SpatialPCA_result$clusterlabel_refine,SpatialPCA_result$location) ### Spatial PCA
```

```{r}
fx_CHAOS_windows(IP_clusterlabel_refine,SpatialPCA_result$location) ### Iterative SVD
```

##### Via Gaussian Mixture
```{r}
fx_CHAOS_windows(pr0,SpatialPCA_result$location) ###PCA
```

```{r}
fx_CHAOS_windows(pr2,SpatialPCA_result$location) ### Spatial PCA
```

```{r}
fx_CHAOS_windows(pr1,SpatialPCA_result$location) ### Iterative SVD
```

### PAS

```{r}
fx_PAS_windows = function(clusterlabel, location){
  # require(parallel)

  matched_location=location
  NAs = which(is.na(clusterlabel))
  if(length(NAs>0)){
    clusterlabel=clusterlabel[-NAs]
    matched_location = matched_location[-NAs,]
  }

  results = mclapply(1:dim(matched_location)[1], fx_kNN, location_in=matched_location,k=10,cluster_in=clusterlabel, mc.cores = 1)
  return(sum(unlist(results))/length(clusterlabel))
}
```


```{r}
fx_PAS_windows(SpatialPCA_result$clusterlabel_refine,SpatialPCA_result$location)
```

```{r}
fx_PAS_windows(IP_clusterlabel_refine,SpatialPCA_result$location)
```

```{r}
fx_PAS_windows(pr0,SpatialPCA_result$location)
```

```{r}
fx_PAS_windows(pr2,SpatialPCA_result$location)
```

```{r}
fx_PAS_windows(pr1,SpatialPCA_result$location)
```

### LISI scores

###SpatialPCA
```{r}
library(lisi)
metadata_DLPFC = data.frame("SpatialPCA" = pr2)
df_test = compute_lisi(SpatialPCA_result$location, metadata_DLPFC, c('SpatialPCA'))
median(df_test$SpatialPCA)
```

### PCA
```{r}
metadata_DLPFC0 = data.frame("PCA" = pr0)
df_test = compute_lisi(SpatialPCA_result$location, metadata_DLPFC0, c('PCA'))
median(df_test$SpatialPCA)
```

### Iterative
```{r}
metadata_DLPFC1 = data.frame("SVD" = pr1)
df_test = compute_lisi(SpatialPCA_result$location, metadata_DLPFC1, c('SVD'))
median(df_test$SpatialPCA)
```





### Plotting leading singular vectors (embeddings, PCs, etc)



```{r}
PCA_denoised0 = PCA_denoised[,-ind_na ]
```



```{r}
par(pty = "s")

plot(PCA_denoised0[1,],PCA_denoised0[2,], col = as.factor(SpatialPCA_result$truth[-ind_na]), main = "exp", xlab = "1st PC", ylab = "2nd PC") ### Iterative SVD
```


```{r}
par(pty = "s")

plot(SpatialPCA_result$SpatialPCs[,-ind_na][1,],SpatialPCA_result$SpatialPCs[,-ind_na][2,], col = as.factor(SpatialPCA_result$truth[-ind_na]), main = "Spatial PCA", xlab = "1st PC", ylab = "2nd PC") 
```

```{r}
par(pty = "s")

plot(t(v0)[1,],t(v0)[2,], col = as.factor(SpatialPCA_result$truth), main = "tau = 0", xlab = "u1", ylab = "u2") 
#plot(v0[,1], v0[,2], col = as.factor(syn.locs$group), main = "PCA")
```



### Visualizations of clusters for different methods
```{r}
library(ggpubr)
cbp=c("#5CB85C" ,"#9C9EDE" ,"#FFDC91", "#4DBBD5" ,"#FF9896" ,"#FED439", "#E377C2", "#FED439")
p1 = plot_cluster(location=SpatialPCA_result$location,as.factor(pr2),pointsize=2,text_size=15 ,title_in=paste0("SpatialPCA"),color_in=cbp)
p2 = plot_cluster(location=SpatialPCA_result$location,SpatialPCA_result$truth,pointsize=2,text_size=15 ,title_in=paste0("Truth"),color_in=cbp)
p3 = plot_cluster(location=SpatialPCA_result$location,as.factor(pr0),pointsize=2,text_size=15 ,title_in=paste0("PCA"),color_in=cbp)
p4 = plot_cluster(location=SpatialPCA_result$location,as.factor(pr1),pointsize=2,text_size=15 ,title_in=paste0("iterative_SVD"),color_in=cbp)

pdf("clusters_Gaussian_Mixture.pdf",width=8, height=8)
ggarrange(p1, p2, p3, p4,
          # labels = c("A", "B", "C"),
          ncol = 2, nrow = 2)
```

```{r}

p1 = plot_cluster(location=SpatialPCA_result$location,SpatialPCA_result$clusterlabel_refine,pointsize=2,text_size=15 ,title_in=paste0("SpatialPCA"),color_in=cbp)
p2 = plot_cluster(location=SpatialPCA_result$location,SpatialPCA_result$truth,pointsize=2,text_size=15 ,title_in=paste0("Truth"),color_in=cbp)
p3 = plot_cluster(location=SpatialPCA_result$location,PCA_clusterlabel_refine,pointsize=2,text_size=15 ,title_in=paste0("PCA"),color_in=cbp)
p4 = plot_cluster(location=SpatialPCA_result$location,IP_clusterlabel_refine,pointsize=2,text_size=15 ,title_in=paste0("iterative_SVD"),color_in=cbp)

pdf("clusters_walktrap.pdf",width=8, height=8)
ggarrange(p1, p2, p3, p4,
          # labels = c("A", "B", "C"),
          ncol = 2, nrow = 2)
```

